---
title: "California water loss data inventory"
output: 
  html_document:
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
    df_print: paged
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(xtable)
library(tidyverse)
library(knitr)
library(ggridges)
library(hrbrthemes)
library(gridExtra)
library(plotly)
library(viridis)

```

This is a summary of some exploratory data analysis examining data from California water audits from 2011-2019

```{r echo = FALSE, cache=TRUE}
ca_data_full <- read_csv('C:/Users/ggorski/OneDrive - DOI/USGS_ML/WaterUse/github/municipal-water-loss-ml/1_data_inventory/out/california_data_all_years.csv',show_col_types = FALSE)


#don't have population data so we can't break it down the same way as Georgia, but we can use length of mains in miles as a
#proxy for the size of the utiliy
data <- ca_data_full %>%
  mutate(ntiles = ntile(`length of mains in miles`,5)) %>%
  group_by(ntiles) %>%
  filter(`non-revenue water as percent by volume of water supplied` > 0) %>%
  mutate(water_losses_pct = (`water losses`/`total water supplied`)*100)
data$ntiles_f <- ordered(data$ntiles, levels = c(1,2,3,4,5), 
                         labels = c('< 67', '67 - 126', '127 - 211','212 - 366','> 367'))

```

# Loss data by length of mains (as a proxy for population)
<br>
<br>
We don't have population data so we can't break it down the same way as Georgia, but we can use length of mains in miles as a
proxy for the size of the utility

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width=12, fig.height=5, cache=TRUE}
plot1 <- ggplot(data, aes(x = `non-revenue water as percent by volume of water supplied`*100, y = ntiles_f, fill = ntiles_f))+
  #geom_density_ridges(rel_min_height = 0.01, scale = 1.35, alpha = 0.8) +
  stat_density_ridges(alpha = 0.8, quantiles = 2, quantile_lines = TRUE)+
  theme_ipsum()+
  xlim(0,80)+
  xlab('Non-revenue water \n (% of Supplied)')+
  ylab('Length of mains in miles')+
  labs(title = 'Non-revenue water 2012-2019')+
  theme(legend.position = 'none', plot.title = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14))

#real water losses
plot2 <- ggplot(data, aes(x = water_losses_pct, y = ntiles_f, fill = ntiles_f))+
  #geom_density_ridges(rel_min_height = 0.01, scale = 1.35, alpha = 0.8) +
  stat_density_ridges(alpha = 0.8, quantiles = 2, quantile_lines = TRUE)+
  theme_ipsum()+
  xlim(0,80)+
  xlab('Real water losses \n (% of Supplied)')+
  ylab('Length of mains in miles')+
  labs(title = 'Water lost 2012-2019')+
  theme(legend.position = 'none', plot.title = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14))
grid.arrange(plot1, plot2, ncol = 2)

```

***
# Water losses through time

Looking at the supply and loss data through time again broken down by length of mains. It doesn't look like there is a clear trend with time. 
<br>
<br>

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width=10, fig.height=10, cache=TRUE}
#water lost through time
#water lost through time
data %>%
  filter(!is.na(`total water supplied`)&!is.na(year)) %>%
  select(year, `total water supplied`, `non-revenue water`, `real losses`, `apparent losses`, ntiles_f) %>%
  mutate(Unbilled.Authorized.Consumption = `non-revenue water`-(`real losses`+`apparent losses`)) %>%
  mutate(Real.Losses.pct = (`real losses`/`total water supplied`), Apparent.Losses.pct = (`apparent losses`/`total water supplied`), 
         Unbilled.Authorized.Consumption.pct = (Unbilled.Authorized.Consumption/`total water supplied`))%>%
  group_by(ntiles_f, year) %>%
  summarise(Real.Losses = median(Real.Losses.pct), Apparent.Losses = median(Apparent.Losses.pct), 
            Unbilled.Authorized.Consumption = median(Unbilled.Authorized.Consumption.pct))%>%
  #filter(Util == 'Adel') %>%
  pivot_longer(cols = Real.Losses:Unbilled.Authorized.Consumption) %>%
  ggplot(aes(x = year, y = value, fill = name, text = name))+
  geom_area()+
  scale_fill_viridis(discrete = TRUE, name = 'Non-Revenue Category')+
  #geom_line(data = supp_data, aes(x = Year, y = Supplied))+
  theme_bw()+
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 14),
        legend.title = element_text(size=14), #change legend title font size
        legend.text = element_text(size=14),
        strip.text.x = element_text(size = 14),
        plot.title = element_text(size=16, face = "bold"))+
  ggtitle('Non-Revenue water grouped by length of mains')+
  ylab('Water Losses as a fraction of water supplied')+
  facet_wrap(.~ntiles_f, ncol = 1, scales = "fixed")

```


***
# Cost of water losses

```{r echo=FALSE, warning = FALSE, message = FALSE, fig.width=12, fig.height=5, cache=TRUE}
#cost of real losses
plot3 <- data %>%
  filter(`annual cost of real losses` >=0) %>%
  mutate(Real.Losses.Cost.Thou = `annual cost of real losses`/1000) %>%
  ggplot(aes(y = ntiles_f, x = Real.Losses.Cost.Thou, fill = ntiles_f, color = ntiles_f))+
  stat_density_ridges(alpha = 0.8, quantiles = 2, quantile_lines = TRUE)+
  theme_ipsum()+
  xlim(0,750)+
  xlab('Cost of real water losses \n (Thousand $)')+
  ylab('Length of mains in miles')+
  labs(title = 'Annual cost of \n real water losses 2012-2019')+
  theme(legend.position = 'none', plot.title = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14))+
  scale_color_manual(values = rep('black',5), guide = "none")

#cost of real losses as percentage of 
plot4 <- data %>%
  filter(`annual cost of real losses` >=0) %>%
  mutate(Real.Losses.Cost.Frac = (`annual cost of real losses`/`total annual operating cost`)*100) %>%
  ggplot(aes(y = ntiles_f, x = Real.Losses.Cost.Frac, fill = ntiles_f, color = ntiles_f))+
  stat_density_ridges(alpha = 0.8, quantiles = 2, quantile_lines = TRUE)+
  theme_ipsum()+
  xlim(0,10)+
  xlab('Cost of real water losses \n (% of operating cost)')+
  ylab('Length of mains in miles')+
  labs(title = 'Annual cost of \n real water losses 2012-2019')+
  theme(legend.position = 'none', plot.title = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14))+
  scale_color_manual(values = rep('black',5), guide = "none")

grid.arrange(plot3, plot4, nrow = 1)
```

***
# Potential drivers of water loss


Crossplots of losses with length of mains, pressure, and number of connections broken down by population served  
<br>

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width=6, fig.height=5, cache=TRUE}
#drivers of loss
g <- data %>%
  filter(`real losses` >=0) %>%
  ggplot(aes(x = `length of mains in miles`, y = `real losses`, fill= ntiles_f, text = paste(`city/town/municipality`,year,sep = ' | ')))+
  geom_point(alpha = 0.6, stroke= 0.25)+
  theme_ipsum()+
  theme(legend.position = 'none')+
  xlab('Length of mains (miles)')+
  ylab('Real losses (million gallons)')+
  labs(title = 'Length of mains')+
  theme(legend.position = 'none', plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        axis.title.x = element_text(size = 13))

ggplotly(g, tooltip = 'text')
```

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width=6, fig.height=5, cache=TRUE}
g1 <- data %>%
  filter(`real losses` >=0) %>%
  ggplot(aes(x = `number active and inactive connections`, y = `real losses`, fill= ntiles_f, text = paste(`city/town/municipality`,year,sep = ' | ')))+
  geom_point(alpha = 0.6, stroke= 0.25)+
  theme_ipsum()+
  theme(legend.position = 'none')+
  xlab('Number of connections')+
  ylab('Real losses (million gallons)')+
  labs(title = 'Number of connections')+
  theme(legend.position = 'none', plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        axis.title.x = element_text(size = 13))

ggplotly(g1, tooltip = 'text')
```

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width=6, fig.height=5, cache=TRUE}
g2 <- data %>%
  filter(`real losses` >=0) %>%
  ggplot(aes(x = `average operating pressure in psi`, y = `real losses`, fill= ntiles_f, text = paste(`city/town/municipality`,year,sep = ' | ')))+
  geom_point(alpha = 0.6, stroke= 0.25)+
  theme_ipsum()+
  theme(legend.position = 'none')+
  xlab('Average operating pressure (PSI)')+
  ylab('Real losses (million gallons)')+
  labs(title = 'System operating pressure')+
  theme(legend.position = 'none', plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        axis.title.x = element_text(size = 13))

ggplotly(g2, tooltip = 'text')


```