---
title: "Georgia Water Loss Data Inventory"
output: 
  html_document:
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
    df_print: paged
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(xtable)
library(tidyverse)
library(knitr)
library(ggridges)
library(hrbrthemes)
library(gridExtra)
library(plotly)
library(viridis)

```


This is a summary of some exploratory data analysis examining data from Georgia water audits from 2011-2019

```{r echo = FALSE, cache=TRUE}
ga_all_yrs <- read_csv('C:/Users/ggorski/OneDrive - DOI/USGS_ML/WaterUse/github/municipal-water-loss-ml/1_data_inventory/out/georgia_all_years.csv',show_col_types = FALSE)
```
# Loss data by population served


A look at the water loss data for all years (2012-2019) for all municipalities broken down by population served  
<br>
<br>

```{r echo=FALSE, results='asis', cache=TRUE}

x <- ga_all_yrs %>%
  filter(!is.na(Population)) %>%
  mutate(ntiles = ntile(Population,5)) %>%
  group_by(ntiles) %>%
  summarise(min_pop = min(Population), max_pop = max(Population), unique_municipalities = n_distinct(WSID), median_supplied = median(Supplied), median_nrl = median(Non.Revenue.Water)) %>%
  mutate(median_nrl_frac = (median_nrl/median_supplied)*100) %>%
  select(!c(median_nrl, ntiles))

#row.names(x) <- c('Least Populous','','','','Most Populous')
kable(x, digits = 1, col.names = c('Minimum population','Maximum population','Unique municipalities','Median supplied water (Mil Gal/Yr)','Median non-revenue water (%)'),
        row.names = FALSE)


```

```{r echo=FALSE, warning = FALSE, message = FALSE, fig.width=12, fig.height=10, cache=TRUE}
data <- ga_all_yrs %>%
  filter(!is.na(Population)) %>%
  mutate(ntiles = ntile(Population,5)) %>%
  group_by(ntiles) %>%
  mutate(Supplied_per_Capita = Supplied/Population) %>%
  mutate(Non_Revenue_water_per_capita = Non.Revenue.Water/Population) %>%
  mutate(Real_Losses_per_capita = Real.Losses/Population) %>%
  mutate(Percent_non_revenue_water = (Non.Revenue.Water/Supplied)*100) %>%
  mutate(Pct_Real_Water_Loss = (Real.Losses/Supplied)*100)

data$ntiles_f <- ordered(data$ntiles, levels = c(1,2,3,4,5), labels = c('< 4,853', '4,855 - 7,062', '7,073 - 13,050','13,260 - 28,025','> 28,1832'))

#Water supplied per capita  
plot1 <- ggplot(data, aes(x = Supplied_per_Capita, y = ntiles_f, fill = ntiles_f))+
  #geom_density_ridges(rel_min_height = 0.01, scale = 1.35, alpha = 0.8) +
  stat_density_ridges(alpha = 0.8, quantiles = 2, quantile_lines = TRUE)+
  theme_ipsum()+
  xlim(0,0.25)+
  xlab('Gallons supplied per capita \n (Million Gallons/Yr/Person)')+
  ylab('Population served')+
  labs(title = 'Water supplied 2012-2019')+
  theme(legend.position = 'none', plot.title = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14))

#non-revenue water per capita
plot2 <- ggplot(data, aes(x = Non_Revenue_water_per_capita, y = ntiles_f, fill = ntiles_f))+
  #geom_density_ridges(rel_min_height = 0.01, scale = 1.35, alpha = 0.8) +
  stat_density_ridges(alpha = 0.8, quantiles = 2, quantile_lines = TRUE)+
  theme_ipsum()+
  xlim(0,0.15)+
  xlab('Non-Revenue water gallons per capita \n (Million Gallons/Yr/Person)')+
  ylab('Population served')+
  labs(title = 'Non-revenue water losses per capita 2012-2019')+
  theme(legend.position = 'none', plot.title = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14))

#Non-Revenue water as a percent of supplied
plot3 <- ggplot(data, aes(x = Percent_non_revenue_water, y = ntiles_f, fill = ntiles_f))+
  #geom_density_ridges(rel_min_height = 0.01, scale = 1.35, alpha = 0.8) +
  stat_density_ridges(alpha = 0.8, quantiles = 2, quantile_lines = TRUE)+
  theme_ipsum()+
  xlim(0,80)+
  xlab('Non-Revenue Water \n (% of Supplied)')+
  ylab('Population served')+
  labs(title = 'Non-revenue water 2012-2019')+
  theme(legend.position = 'none', plot.title = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14))

#Real water loss as a percet of supplied
plot4 <- ggplot(data, aes(x = Pct_Real_Water_Loss, y = ntiles_f, fill = ntiles_f))+
  #geom_density_ridges(rel_min_height = 0.01, scale = 1.35, alpha = 0.8) +
  stat_density_ridges(alpha = 0.8, quantiles = 2, quantile_lines = TRUE)+
  theme_ipsum()+
  xlim(0,80)+
  xlab('Real water losses \n (% of Supplied)')+
  ylab('Population served')+
  labs(title = 'Water lost 2012-2019')+
  theme(legend.position = 'none', plot.title = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14))


grid.arrange(plot1, plot2, plot3, plot4, ncol = 2)
```

***
# Water losses through time

Looking at the supply and loss data through time, hover over the dots to view the municipality. It doesn't look like there is a clear trend with time.  
<br>
<br>

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width=10, fig.height=10, cache=TRUE}
p <- ga_all_yrs %>%
  filter(!is.na(Population)) %>%
  mutate(ntiles = ntile(Population,5)) %>%
  mutate(ntiles_f = ordered(ntiles, levels = c(1,2,3,4,5), 
                          labels = c('Population < 4,853', 'Population 4,855 - 7,062', 'Population 7,073 - 13,050',
                                     'Population 13,260 - 28,025','Population > 28,1832'))) %>%
  select(Year, Util, Population, Supplied, Non.Revenue.Water, Real.Losses, Apparent.Losses, ntiles_f) %>%
  mutate(Unbilled.Authorized.Consumption = Non.Revenue.Water-(Real.Losses+Apparent.Losses)) %>%
  mutate(Real.Losses.pct = (Real.Losses/Supplied), Apparent.Losses.pct = (Apparent.Losses/Supplied), 
         Unbilled.Authorized.Consumption.pct = (Unbilled.Authorized.Consumption/Supplied))%>%
  group_by(ntiles_f, Year) %>%
  summarise(Real.Losses = median(Real.Losses.pct), Apparent.Losses = median(Apparent.Losses.pct), 
            Unbilled.Authorized.Consumption = median(Unbilled.Authorized.Consumption.pct))%>%
  #filter(Util == 'Adel') %>%
  pivot_longer(cols = Real.Losses:Unbilled.Authorized.Consumption) %>%
  ggplot(aes(x = Year, y = value, fill = name, text = name))+
  geom_area()+
  scale_fill_viridis(discrete = TRUE, name = 'Non-Revenue Category')+
  #geom_line(data = supp_data, aes(x = Year, y = Supplied))+
  theme_bw()+
  theme(axis.text = element_text(size = 14),
            axis.title = element_text(size = 14),
            legend.title = element_text(size=14), #change legend title font size
            legend.text = element_text(size=14),
           strip.text.x = element_text(size = 14),
            plot.title = element_text(size=16, face = "bold"))+
  ggtitle('Non-Revenue Water by Municipality Population')+
  ylab('Water Losses as a fraction of water supplied')+
  facet_wrap(.~ntiles_f, ncol = 1, scales = "fixed")
p
```

***
# Cost of water losses

```{r echo=FALSE, warning = FALSE, message = FALSE, fig.width=12, fig.height=5, cache=TRUE}

plot5 <- ga_all_yrs %>%
  filter(Real.Losses.Cost >=0) %>%
  filter(!is.na(Population)&!is.na(Real.Losses.Cost)) %>%
  mutate(ntiles = ntile(Population,5)) %>%
  mutate(ntiles_f = ordered(ntiles, levels = c(1,2,3,4,5), 
                            labels = c('Population < 4,853', 'Population 4,855 - 7,062', 'Population 7,073 - 13,050',
                                       'Population 13,260 - 28,025','Population > 28,1832'))) %>%
  mutate(Real.Losses.Cost.Thou = Real.Losses.Cost/1000) %>%
  ggplot(aes(y = ntiles_f, x = Real.Losses.Cost.Thou, fill = ntiles_f, color = ntiles_f))+
  stat_density_ridges(alpha = 0.8, quantiles = 2, quantile_lines = TRUE)+
  theme_ipsum()+
  xlim(0,500)+
  xlab('Cost of real water losses \n (Thousand $)')+
  ylab('Population served')+
  labs(title = 'Annual cost of \n real water losses 2012-2019')+
  theme(legend.position = 'none', plot.title = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14))+
  scale_color_manual(values = rep('black',5), guide = "none") #+
  #scale_fill_manual(values = my_pal, guide = "none")

plot6 <- ga_all_yrs %>%
  filter(Real.Losses.Cost >=0) %>%
  filter(!is.na(Population)&!is.na(Real.Losses.Cost)&!is.na(Operating.Cost)) %>%
  mutate(ntiles = ntile(Population,5)) %>%
  mutate(ntiles_f = ordered(ntiles, levels = c(1,2,3,4,5), 
                            labels = c('Population < 4,853', 'Population 4,855 - 7,062', 'Population 7,073 - 13,050',
                                       'Population 13,260 - 28,025','Population > 28,1832'))) %>%
  mutate(Real.Losses.Cost.Frac = (Real.Losses.Cost/Operating.Cost)*100) %>%
  ggplot(aes(y = ntiles_f, x = Real.Losses.Cost.Frac, fill = ntiles_f, color = ntiles_f))+
  stat_density_ridges(alpha = 0.8, quantiles = 2, quantile_lines = TRUE)+
  theme_ipsum()+
  xlim(0,30)+
  xlab('Cost of real water losses \n (% of operating cost)')+
  ylab('Population served')+
  labs(title = 'Annual cost of \n real water losses 2012-2019')+
  theme(legend.position = 'none', plot.title = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14))+
  scale_color_manual(values = rep('black',5), guide = "none")#+
  #scale_fill_manual(values = my_pal, guide = "none")

grid.arrange(plot5, plot6, nrow = 1)
```

***
# Potential drivers of water loss


Crossplots of losses with length of mains, pressure, and number of connections broken down by population served  
<br>

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width=10, fig.height=5, cache=TRUE}
g <- data %>%
  filter(Real.Losses >=0) %>%
  ggplot(aes(x = Length.of.Mains, y = Real.Losses, fill= ntiles_f, text = paste(Util,Year,sep = ' | ')))+
  geom_point(alpha = 0.6, stroke= 0.25)+
  facet_wrap(~ntiles_f, scales = c('free'), nrow = 1)+
  theme_ipsum()+
  theme(legend.position = 'none')+
  xlab('Length of Mains (Miles)')+
  ylab('Real Losses (Million Gallons)')+
  labs(title = 'Length of Mains')+
    theme(legend.position = 'none', plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        axis.title.x = element_text(size = 13))
  
ggplotly(g, tooltip = 'text')
```

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width=10, fig.height=5, cache=TRUE}
g <- data %>%
  filter(Real.Losses >=0) %>%
  ggplot(aes(x = Service.COnnections, y = Real.Losses, fill= ntiles_f, text = paste(Util,Year,sep = ' | ')))+
  geom_point(alpha = 0.6, stroke= 0.25)+
  facet_wrap(~ntiles_f, scales = c('free'), nrow = 1)+
  theme_ipsum()+
  theme(legend.position = 'none')+
  xlab('Service Connections')+
  ylab('Real Losses (Million Gallons)')+
  labs(title = 'Number of Service Connections')+
    theme(legend.position = 'none', plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        axis.title.x = element_text(size = 13))
  
ggplotly(g, tooltip = 'text')
```


```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width=10, fig.height=5, cache=TRUE}
g <- data %>%
  filter(Real.Losses >=0) %>%
  ggplot(aes(x = Pressure, y = Real.Losses, fill= ntiles_f, text = paste(Util,Year,sep = ' | ')))+
  geom_point(alpha = 0.6, stroke= 0.25)+
  facet_wrap(~ntiles_f, scales = c('free'), nrow = 1)+
  theme_ipsum()+
  theme(legend.position = 'none')+
  xlab('Average Operating Pressure (PSI)')+
  ylab('Real Losses (Million Gallons)')+
  labs(title = 'System Pressure')+
    theme(legend.position = 'none', plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        axis.title.x = element_text(size = 13))
  
ggplotly(g, tooltip = 'text')
```

