---
title: "Georgia Data Inventory"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(xtable)
library(tidyverse)
library(knitr)
library(ggridges)
library(hrbrthemes)
library(gridExtra)
library(plotly)
```

```{r echo = FALSE}
ga_all_yrs <- read_csv('C:/Users/ggorski/OneDrive - DOI/USGS_ML/WaterUse/github/municipal-water-loss-ml/1_data_inventory/out/georgia_all_years.csv',show_col_types = FALSE)
```
##### A look at the water loss data for all years (2012-2019) for all municipalities broken down by population served  
<br>
<br>

```{r echo=FALSE, results='asis'}

x <- ga_all_yrs %>%
  filter(!is.na(Population)) %>%
  mutate(ntiles = ntile(Population,5)) %>%
  group_by(ntiles) %>%
  summarise(min_pop = min(Population), max_pop = max(Population), unique_municipalities = n_distinct(WSID), mean_supplied = mean(Supplied), mean_lost = mean(Water.Losses)) %>%
  mutate(mean_lost_frac = (mean_lost/mean_supplied)*100) %>%
  select(!c(mean_lost, ntiles))

#row.names(x) <- c('Least Populous','','','','Most Populous')
kable(x, digits = 1, col.names = c('Minimum Population','Maximum Population','Unique Municipalities','Mean Supplied Water (Mil Gal/Yr)','Mean Water Lost (%)'),
        row.names = FALSE)


```

```{r echo=FALSE, warning = FALSE, message = FALSE, fig.width=12, fig.height=4}
data <- ga_all_yrs %>%
  filter(!is.na(Population)) %>%
  mutate(ntiles = ntile(Population,5)) %>%
  group_by(ntiles) %>%
  mutate(Supplied_per_Capita = Supplied/Population) %>%
  mutate(Percent_water_loss = (Water.Losses/Supplied)*100)

data$ntiles_f <- ordered(data$ntiles, levels = c(1,2,3,4,5), labels = c('< 4,853', '4,855 - 7,062', '7,073 - 13,050','13,260 - 28,025','> 28,1832'))
  
plot1 <- ggplot(data, aes(x = Supplied_per_Capita, y = ntiles_f, fill = ntiles_f))+
  geom_density_ridges(rel_min_height = 0.005, scale = 0.75) +
  theme_ipsum()+
  xlim(0,0.25)+
  xlab('Gallons supplied per capita (Million Gallons/Yr/Person)')+
  ylab('Population Served')+
  labs(title = 'Water supplied 2012-2019')+
  theme(legend.position = 'none', plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        axis.title.x = element_text(size = 13))


plot2 <- ggplot(data, aes(x = Percent_water_loss, y = ntiles_f, fill = ntiles_f))+
  geom_density_ridges(rel_min_height = 0.01, scale = 1.35) +
  theme_ipsum()+
  xlim(0,80)+
  xlab('Water Lost (% of Supplied)')+
  ylab('Population Served')+
  labs(title = 'Water lost 2012-2019')+
  theme(legend.position = 'none', plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        axis.title.x = element_text(size = 13))

grid.arrange(plot1, plot2, ncol = 2)
```

***

##### Looking at the supply and loss data through time, hover over the dots to view the municipality. It doesn't look like there is a clear trend with time, but maybe the outliers are interesting?  
<br>
<br>

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width=10, fig.height=4}
t <- data %>%
  filter(ntiles == 5) %>%
  group_by(Year) %>%
  pivot_longer(cols = c('Supplied','Water.Losses')) %>%
  select(Year, Util, ntiles, name, value)

plot_ly(t, x = ~Year, y = ~value, color = ~name, type = 'box', text = ~Util,colors = c('#0C4B8E','#BF382A'))%>% 
  layout(boxmode = "group", 
         xaxis = list(title='Year'), 
         yaxis = list(title='Water (Million Gallons)'),
         title = 'Water Supplied Municipalities Population > 28,183')
```
```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width=10, fig.height=4}
t <- data %>%
  filter(ntiles == 4) %>%
  group_by(Year) %>%
  pivot_longer(cols = c('Supplied','Water.Losses')) %>%
  select(Year, Util, ntiles, name, value)

plot_ly(t, x = ~Year, y = ~value, color = ~name, type = 'box', text = ~Util,colors = c('#0C4B8E','#BF382A'))%>% 
  layout(boxmode = "group", 
         xaxis = list(title='Year'), 
         yaxis = list(title='Water (Million Gallons)'),
         title = 'Water Supplied Municipalities Population 13,260 - 28,025')
```

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width=10, fig.height=4}
t <- data %>%
  filter(ntiles == 3) %>%
  group_by(Year) %>%
  pivot_longer(cols = c('Supplied','Water.Losses')) %>%
  select(Year, Util, ntiles, name, value)

plot_ly(t, x = ~Year, y = ~value, color = ~name, type = 'box', text = ~Util,colors = c('#0C4B8E','#BF382A'))%>% 
  layout(boxmode = "group", 
         xaxis = list(title='Year'), 
         yaxis = list(title='Water (Million Gallons)'),
         title = 'Water Supplied Municipalities Population 7,073 - 13,050')
```

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width=10, fig.height=4}
t <- data %>%
  filter(ntiles == 2) %>%
  group_by(Year) %>%
  pivot_longer(cols = c('Supplied','Water.Losses')) %>%
  select(Year, Util, ntiles, name, value)

plot_ly(t, x = ~Year, y = ~value, color = ~name, type = 'box', text = ~Util,colors = c('#0C4B8E','#BF382A'))%>% 
  layout(boxmode = "group", 
         xaxis = list(title='Year'), 
         yaxis = list(title='Water (Million Gallons)'),
         title = 'Water Supplied Municipalities Population 4,855 - 7,062')
```

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width=10, fig.height=4}
t <- data %>%
  filter(ntiles == 1) %>%
  group_by(Year) %>%
  pivot_longer(cols = c('Supplied','Water.Losses')) %>%
  select(Year, Util, ntiles, name, value)

plot_ly(t, x = ~Year, y = ~value, color = ~name, type = 'box', text = ~Util,colors = c('#0C4B8E','#BF382A'))%>% 
  layout(boxmode = "group", 
         xaxis = list(title='Year'), 
         yaxis = list(title='Water (Million Gallons)'),
         title = 'Water Supplied Municipalities Population < 4,853')
```

***

##### Crossplots of losses with length of mains, pressure, and number of connections broken down by population served  
<br>

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width=10, fig.height=5}
g <- data %>%
  ggplot(aes(x = Length.of.Mains, y = Real.Losses, fill= ntiles_f, text = paste(Util,Year,sep = ' | ')))+
  geom_point(alpha = 0.6, stroke= 0.25)+
  facet_wrap(~ntiles_f, scales = c('free'), nrow = 1)+
  theme_ipsum()+
  theme(legend.position = 'none')+
  xlab('Length of Mains (Miles)')+
  ylab('Real Losses (Million Gallons)')+
  labs(title = 'Length of Mains')+
    theme(legend.position = 'none', plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        axis.title.x = element_text(size = 13))
  
ggplotly(g, tooltip = 'text')
```

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width=10, fig.height=5}
g <- data %>%
  ggplot(aes(x = Service.COnnections, y = Real.Losses, fill= ntiles_f, text = paste(Util,Year,sep = ' | ')))+
  geom_point(alpha = 0.6, stroke= 0.25)+
  facet_wrap(~ntiles_f, scales = c('free'), nrow = 1)+
  theme_ipsum()+
  theme(legend.position = 'none')+
  xlab('Service Connections')+
  ylab('Real Losses (Million Gallons)')+
  labs(title = 'Number of Service Connections')+
    theme(legend.position = 'none', plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        axis.title.x = element_text(size = 13))
  
ggplotly(g, tooltip = 'text')
```


```{r echo = FALSE, warning = FALSE, message = FALSE, fig.width=10, fig.height=5}
g <- data %>%
  ggplot(aes(x = Pressure, y = Real.Losses, fill= ntiles_f, text = paste(Util,Year,sep = ' | ')))+
  geom_point(alpha = 0.6, stroke= 0.25)+
  facet_wrap(~ntiles_f, scales = c('free'), nrow = 1)+
  theme_ipsum()+
  theme(legend.position = 'none')+
  xlab('Average Operating Pressure (PSI)')+
  ylab('Real Losses (Million Gallons)')+
  labs(title = 'System Pressure')+
    theme(legend.position = 'none', plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        axis.title.x = element_text(size = 13))
  
ggplotly(g, tooltip = 'text')
```

